---
title: "Introduction to R: Lecture 5"
subtitle: "Topics: Data Analysis and Graphs"
author: "Sabrina Nardin, Summer 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---

## Agenda {.center}

::: {.agenda-list}
1. The `scorecard` Dataset  
2. Match Graph Types to Variable Types  
3. Practice  
:::

---

## Our Goals Today

### Practice using graphs for data analysis. Specifically:

- **Display variation and co-variation:** Learn how to visualize the distribution of a single variable (e.g., `cost` to attend a school) and how two or more variables vary togheter (e.g., `cost` and `admrate`).

- **Match graph type to variable type:** Choose appropriate graph types depending on whether variables are categorical (e.g., school `type`), continuous (e.g., `cost`), or both.

- **Interpret the graph:** Practice describing what a graph reveals — such as trends, group differences, and outliers.


# 1. The scorecard Dataset {.slide .center .middle}

---

## About the scorecard dataset

The U.S. Department of Education collects annual statistics on colleges and universities in the United States: <https://collegescorecard.ed.gov/data>

This dataset includes variables such as: 

- `name`: name of the school 
- `state`: state where the school is located  
- `type`: school type (e.g., Public, Private Nonprofit, Private For-Profit)  
- `admrate`: admission rate (e.g., 0.91 = 91%)  
- `cost`: published cost of attendance  
- `netcost`: net cost of attendance after financial aid  
- `satavg`: average SAT score of admitted students  
- `...`: many additional variables

We’ll focus on a subset of this data from the 2018–2019 academic year.

---

## About the scorecard dataset

```r
library(tidyverse)
library(rcis)
data(scorecard)
glimpse(scorecard)
```

```{r}
library(tidyverse)
library(rcis)
data(scorecard)
glimpse(scorecard)
```

---

## The scorecard dataset: making a plot

#### Which type of college has the highest average SAT score?

```r
# with faceted histogram
ggplot(data = scorecard, mapping = aes(x = satavg)) +
  geom_histogram() + 
  facet_wrap(facets = vars(type))
```

```{r}
# with faceted histogram
ggplot(data = scorecard, mapping = aes(x = satavg)) +
  geom_histogram() + 
  facet_wrap(facets = vars(type))
```

#### [What's a histogram?](https://datatab.net/tutorial/charts){target="_blank"}

---

#### Which type of college has the highest average SAT score?

```r
# with a boxplot
ggplot(data = scorecard, mapping = aes(x = type, y = satavg)) +
  geom_boxplot()
```

```{r}
# with a boxplot
ggplot(data = scorecard, mapping = aes(x = type, y = satavg)) +
  geom_boxplot()
```

#### [What's a boxplot?](https://datatab.net/tutorial/box-plot){target="_blank"}

---

## The scorecard dataset: interpreting a plot

#### What do these graphs reveal about average SAT scores by type of college?

::: {.callout-note title="Interpreting plots is as important as writing code"}

According to these graphs, **private nonprofit schools** have the highest average SAT scores, closely followed by **public schools**, and then **private for-profit schools**.

But this interpretation doesn't tell the full story:

- From the histogram, we can see that each school type includes a different number of colleges. i.e., they have **different sample sizes** (private for-profit schools have far fewer schools than others).

- This matters because averages based on small groups may not represent the broader category well. Looking at the full distribution gives us a more complete understanding.
:::

---

## The scorecard dataset: asking more questions, answering them with dplyr

How many schools are in **each type?** 
```r
scorecard %>% count(type) 
```

Which schools are categorized as **private for-profit?**
```r
scorecard %>%
  filter(type == "Private, for-profit") %>%
  select(name, state, type, satavg, cost)
```

What about the **University of Chicago?**
```r
scorecard %>%
  filter(scorecard, name == "University of Chicago")
```

Etc. 

<!--
The code below creates a summary dataset with `type` of university as rows and the `satavg` mean as a column:

```{r}
scorecard %>% 
  group_by(type) %>%
  summarize(mean_type = mean(satavg, na.rm = TRUE))
```
-->

---

## Once you draw a plot, ask yourself:

::: columns
::: column

#### Substantive questions:
- What does this graph tell?
- Are there patterns? Outliers?
- What hypotheses can I generate?
- What else I want to know?
- Should I dig deeper with `dplyr`?
- Etc.

:::
::: column

#### Stylistic questions:
- Is the chosen plot appropriate (match variable type)?
- Is the plot clear and easy to understand?
- Is it too busy or too simple?
- Could it be improved? (titles, labels, colors, etc.)
- Does I need to manipulate the data first?
- Etc.

:::
:::

# 2. Matching Graph Types to Variable Types {.slide .center .middle}

---

## Types of Visualizations and Best Graph Types

Do I want to represent variation in:

- **A single variable?**
- **Two variables?**
- **Three variables?**

What type(s) are my variables?

- **Continuous** (e.g., `satavg` average SAT score, `cost` published cost)
- **Categorical** (e.g., `type` school type, `state` U.S. state)
- Other types (often treated as categorical): ordinal, nominal, binary

::: {.callout-tip}
Think about which variables you want to display and their type before choosing a graph!
:::

---

## Univariate (One Variable)

To show **how values vary within a single variable**:

- One **continuous** variable → **Histogram**
- One **categorical** variable → **Bar Chart**

---

## Univariate (One Variable) 

::: {.smaller}

| Function           | What Gets Counted                                 | When to Use             |
|--------------------|---------------------------------------------------|--------------------------|
| `geom_histogram()` | How many values fall into each numeric bin        | When `x` is continuous   |
| `geom_bar()`       | How many observations are in each category        | When `x` is categorical  |

:::

---

## Bivariate (Two Variables)

To show **how two variables co-vary**:

- Two **continuous** variables → **Scatterplot**
- One **categorical** + one **continuous** → **Box Plot**
- Two **categorical** variables → **Grouped or Stacked Bar Chart**

---

## Multivariate (Three Variables)

To compare patterns **across subgroups**:

- One **categorical** + two **continuous** → **Faceted Scatterplot**
- Two **categorical** + one **continuous** → **Grouped Box Plot**

::: tip
These are the most common combinations. There are more options as you explore deeper!
:::

---

## Main Graph Types and When to Use Them

| Type of Graph        | `ggplot2` Function                                                              | Input Variables                                                                 | Goal                                  |
|----------------------|----------------------------------------------------------------------------------------|----------------------------------------------------------------------------------|----------------------------------------------------------|
| Histogram            | [geom_histogram()](https://ggplot2.tidyverse.org/reference/geom_histogram.html)       | One **continuous** (e.g., age, income)                              | Show the **distribution** of values                     |
| Bar Chart            | [geom_bar()](https://ggplot2.tidyverse.org/reference/geom_bar.html)                   | One **categorical** (e.g., region, gender)                           | Show **frequencies** or **counts** of categories        |
| Stacked Bar Chart    | [geom_bar() + fill](https://ggplot2.tidyverse.org/reference/geom_bar.html)            | Two **categorical** variables (e.g., region by gender)                         | Compare **parts of a whole** across categories          |
| Scatterplot          | [geom_point()](https://ggplot2.tidyverse.org/reference/geom_point.html)               | Two **continuous** variables (e.g., height vs. weight, price vs. rating)        | Show **relationship** or **correlation**                |
| Box Plot             | [geom_boxplot()](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)           | One **continuous** + one **categorical** (e.g., income by gender)           | Compare **distributions**, spot **outliers**            |
| Faceted Scatterplot  | [facet_wrap() + geom_point()](https://ggplot2.tidyverse.org/reference/facet_wrap.html) | Two **continuous** + one **categorical** (e.g., by country or year)             | Compare **patterns across groups**                      |


# 3. Practice {.slide .center .middle}

---

## 💻 Practice

On the next slide, you’ll see a set of tasks. In small groups, use the scorecard dataset to create the most appropriate graph for each one.

- **Before plotting:** Consider the type of variable and the type of variation you need to represent. Use the slides as reference.

- **While plotting:** Keep it simple, as you would for an initial Exploratory Data Analysis (e.g., no need to add labels, legends, color adjustments, scales, themes, facets, etc.)

- **After plotting:** Stare at the graph... look for patterns, outliers, or any notable features, and substantively interpret the graph.

---

## 💻 Practice

<!-- For regular CSS class, add some dpyr commands in task and also bring back TASK 4  Display the total number of schools in the U.S. by school type (n = 3) and by state (n = 54). *Note: the initial graph you generate here may lack visual appeal. Focus on identifying potential improvements rather than implementing them for now.* -->

Share your solutions here <https://codeshare.io/vAzK44>

TASK 1: Create a plot that displays the annual total cost of school attendance across the U.S.. (Hint: try `geom_histogram()` with the variable `cost`)

TASK 2: Create a plot that displays the total number of schools in the U.S. by school type (Hint: try `geom_bar()` with the variable variable `type`)

TASK 3: Create a plot that displays the annual total cost and net cost of attendance to schools in the U.S. (variables `cost` and `netcost`)

TASK 4: Create a plot that displays the annual total cost of attendance by school type (variables `cost` and `type`)

TASK 5: Create a plot that displays the annual total cost of attendance and net cost of attendance by school type (variables `cost`, `netcost`, `type`)

---

## Takeaways and Tips

Follow this approach to move from quick exploration to polished, communicative plots:

::: columns
::: column

#### When exploring your data:
- Start with simple and plots to get a sense of the distribution and relationships
- Focus on interpretation first, not style
- Ask questions: What stands out? Are there outliers, trends, or surprises? What else do you want to know?

:::
::: column

#### Once you settle on a plot:
- Refine the code to improve readability and aesthetics
- Add styling elements (labels, scales, legends, and themes. etc. as needed)
- Use polished plots for assignments, reports, or presentations

:::
:::

*Typically, researchers create many exploratory plots and only a few make it to the final report.*

---

## 💻 Download Today Materials

Download today’s class materials from our website for: 

- further insights into these tasks
- additional practice exercises

---

## Recap: What We Learned Today

- Pick graph types that match the types of variables in your data
- Read and interpret visualizations
- Pracice using `ggplot` and `dplyr`

---

## To print these slides as pdf

Click on the icon bottom-right corner \> Tools \> PDF Export Mode \> Print as a Pdf

