---
title: "Introduction to R: Lecture 9"
subtitle: "Topics: Conditional Statements with if, else, and else if"
author: "Sabrina Nardin, Summer 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---

# Agenda {.center}

::: {.agenda-list}
1. Overview of Control Structures: Conditional Statements and Loops
2. Deeper Dive into Conditional Statements
:::

```{r, echo=FALSE, results='asis'}
cat('<span style="font-size: 0.8em; color: #666;">These slides were last updated on', format(Sys.Date(), '%B %d, %Y'), '</span>')
```

<!--
<span style="font-size: 0.8em; color: #666;">
*These slides were last updated on **`r format(Sys.Date(), '%B %d, %Y')`***
</span>
-->

# 1. Overview of Control Structures {.slide .center .middle}

---

## Control Structures: What's the Idea?

Up to this point, our R code has been running line by line following in a fixed sequence.

<br>

::: {.callout-tip title="**Now we ask: Can our code be dynamic vs fixed?**"}

For Example: Can it skip some lines and run others depending on a condition? Or can it repeat itself when needed?

Yes, with control structures!
:::

</br>

---

## What Are Control Structures?

Control structures let us write code that can:

- **Make decisions**  
  → with **conditional statements** (`if`, `if...else`, etc.)

- **Repeat actions**  
  → with **loops** (`for`, `while`)

Conditional statements and loops — that is, control structures — let us **control the flow of execution**.  They allow our code to **respond dynamically to different situations**, rather than always running in the same fixed order like we’ve done so far.

<!-- 
Make decisions: the code uses conditions, evaluates them, and chooses what to do  
Repeat actions: the code runs certain lines multiple times when needed  

*Today: Conditional Statements and Next time: Loops*
-->

# 2. Deeper Dive into Conditional Statements  {.slide .center .middle}

---

## Conditional Statements

**Conditional statements let our code make decisions: the code checks conditions, evaluates them, and chooses what to do.**

Said more formally, we use logical tests to determine which code should run:

1. Single test with `if... else`

2. Multiple tests with `if... else if... else`

3. Nested tests by placing one `if` inside another

4. Vectorized tests with `ifelse()` (or `if_else()` in tidyverse) to apply a condition to each element of a vector

<!--
## Loops

**Loops let our code repeat actions by running certain lines multiple times when needed**

1. For Loop

2. While Loops

-->

---

## 💻 Practice

1. Download today's class materials from the course website.

2. Open the file called `warm-up.R`.

3. Before running the code: look over the syntax and try to predict what the code will output

4. Run the code and compare it to your prediction.

5. Be ready to share at least one question about what the code.

---

## 1. Single test with if... else

Use `if...else` when your code needs to test **one condition** and take one of two possible actions, depending on whether the condition is `TRUE` or `FALSE`.

Syntax:
```r
if (condition) {
    # action performed when condition is TRUE
    action1
} else {
    # action performed when condition is FALSE
    action2
}
```

Example:
```{r echo = TRUE, results = "hide"}
age <- 14

if (age > 16) {
  print("You can get a driving license")
} else {
  print("You cannot drive")
}
```

<!-- NOTE IN CLASS
Examine the syntax and the curly braces and indentation
-->

---

## 1. Single test with if... else

Another Example: What's the output of this code?

```{r, echo = TRUE, results = "hide"}
movie_rating <- c(4.5, 4.2, 5.0, 3.8, 2.9, 1.7)
movie_avg_rating <- mean(movie_rating)

if (movie_avg_rating >= 4) {
  print("This movie is great!")
} else {
  print("Not a favorite.")
}
```

---

## 2. Multiple tests with if... else if... else

Use `if...else if...else` when your code needs to test **multiple conditions** and take different actions depending on which condition is `TRUE`.

Syntax:
```r
if (condition1) {
    # action performed when condition 1 is TRUE
    action1
} else if (condition2) {
    # action performed when condition 2 is TRUE
    action2
} else {
    # action performed when conditions 1 and 2 are FALSE
    action3
}
```
<!-- IN CLASS
Note all we do is adding one or more **`else if`** statement(s) between the initial `if` and the final `else`
-->

Example:

```{r, echo = TRUE, results = "hide"}
x <- 0

if (x > 0) {
  print("x is positive")
} else if (x < 0) {
  print("x is negative")
} else {
  print("x is zero")
}
```

---

## Note: with conditional statements sequence matters!

Conditional statements are evaluated **in order** so the sequence of your code matters:

- R checks each condition **one by one**, from top to bottom  
- As soon as it finds a condition that is `TRUE`, it **runs that block** and **skips the rest**

---

## 2. Multiple tests with if... else if... else

Another Example with multiple `else if` statements:

```r
# Take user input
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

# Determine weather based on temperature
if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
} else {
  weather <- "Freezing"
}
print(weather)
```

---

## 💻 Practice

1. What happens if the conditions are out of order? Move `temperature >= 10` before `temperature >= 20` in the code and run it with temperature 25

<!--
```r
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 10) {   # moved up
  weather <- "Breezy"
} else if (temperature >= 20) {
  weather <- "Cool"
} else {
  weather <- "Freezing"
}
print(weather)
```

Problem: If temperature is 25, R hits temperature >= 10 first (which is TRUE) and never reaches temperature >= 20. So it assigns "Breezy" instead of "Cool".
-->

2. Add another `else if` statement to further refine the code:
If the temperature is between 0 and 10 (inclusive of 0, exclusive of 10), set the weather to "Cold". Hint: use the `&` operator to combine two conditions.

<!--
```r
# Take user input
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

# Determine weather based on temperature
if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
} else if (temperature >= 0 & temperature < 10) {  
  weather <- "Cold"
} else {
  weather <- "Freezing"
}
print(weather)
```

Takeaway: Use & when both conditions must be TRUE (here, >= 0 and < 10). If both are TRUE, then the code inside the {} runs

-->
 
Once done, [share your code here](https://codeshare.io/29X3QM){target="_blank"}.
 
---

## 3. Nested tests by placing one if inside another

<!-- The %% modulo operator calculates the remainder from a division operation. Here 15 : 2 = 7, 1 is the reminder, thus the number is odd -->

Use nested `if...else` statements when your code needs to check **conditions within conditions**. This allows you to make a second decision **only after** a first condition is met.

Example (note the `%%` returns the remainder after division):

```{r, echo = TRUE, results = "hide"}
x <- 15

if (x > 0) {
  if (x %% 2 == 0) {
    print("x is a positive even number")
  } else {
    print("x is a positive odd number")
  }
} else {
  if (x %% 2 == 0) {
    print("x is a negative even number")
  } else {
    print("x is a negative odd number")
  }
}
```

<!-- ASK IN CLASS
- What output do you expect if `x <- -6`?
- What about if `x <- 0
-->

---

## 3. Nested tests by placing one if inside another

Same example, but this checks also the condition `x <- 0`:

```{r, echo = TRUE, results = "hide"}
x <- 0

if (x > 0) {
  if (x %% 2 == 0) {
    print("x is a positive even number")
  } else {
    print("x is a positive odd number")
  }
} else if (x < 0) {
  if (x %% 2 == 0) {
    print("x is a negative even number")
  } else {
    print("x is a negative odd number")
  }
} else {
  print("x is zero")
}

```

<!-- NOTES IN CLASS
What will the output be for:
- x <- 4
- x <- -3
- x <- 0
-->

---

## 4. Vectorized tests with ifelse() 

Use `ifelse()` to apply a condition to **each element of a vector**. This is called a "vectorized test", and fits well R's vector-based design.

Syntax:
```
ifelse (condition to be evaluated,
    action performed when condition is TRUE,
    action performed when condition is FALSE)
```

Example:
```{r, echo = TRUE, results = "hide"}
y <- 3

ifelse(sqrt(16) > y, 
       sqrt(16),
       0)
```

<!--
if-else: more flexibile and handles complex conditional logic; returns a scalar

ifelse(): simpler and more suited for simple assignments or to transform variables in a dataframe; returns a vector
-->

---

## 4. Vectorized tests with ifelse() 

Another example: What is the output of this code?

```{r, echo = TRUE, results = "hide"}
numbers <- c(10, 6, 7)
ifelse(numbers %% 2 == 1, 
       "odd",
       "even")
```

<br>

::: {.callout-note title="How `ifelse()` works:"}

- The input to `ifelse()` is often a **vector**, not just a single value, so the output is also a **vector**. The code evaluates **each element of the vector**, applying the specified action for each case.

- This works because **`ifelse()` supports vectorized operations**: operations directly applied on entire vectors, rather than looping through individual elements one-by-one.

:::

<!--
How can we rewrite the same code using the standard `if-else` statement?

```
# won't work
numbers <- c(10, 6, 7)
if (numbers %% 2 == 1) { 
  print("odd")
} else {
  print("even")
}
```

```
# need a loop
numbers <- c(10, 6, 7)
for (i in 1:length(numbers)) {
  if (numbers[i] %% 2 == 1) {
    print("odd")
    } 
else 
  print("even")
 }
```
-->

---

## 4. Vectorized tests with ifelse() 

Example: What is the output of this code?

```{r, echo = TRUE, results = "hide"}
library(tidyverse)

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

# using base R to access column Scores
ifelse(qualify$"Scores" > 30, 
       "Admitted", 
       "Rejected")

# using dplyr
qualify %>%
  mutate(status = ifelse(Scores > 30,
                       "Admitted",
                       "Rejected"))
```


---

## 4. Vectorized tests with ifelse() 

In the previous example, we only had two possible categories: "Admitted" and "Rejected"

What happens to our code if we add a third one, say "Waitlisted"? We need to use nested `ifelse()` statements:

<!--
```{r, echo = TRUE, results = "hide"}
library(tidyverse)

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

ifelse(qualify$Scores > 35,  # condition to be evaluated
       "Admitted",           # if true, assign "Admitted"
       ifelse(qualify$Scores >= 30,  # if false, evaluate this other condition
              "Waitlisted",          # if true, assign "Waitlisted"
              "Rejected"))           # if false, assign "Rejected"

```
-->

```{r, echo = TRUE, results = "hide"}

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

qualify %>%
  mutate(Status = ifelse(Scores > 35,            # first condition
                         "Admitted",             # if TRUE
                         ifelse(Scores >= 30,    # second condition (only if first is FALSE)
                                "Waitlisted",    # if TRUE
                                "Rejected")))    # if both FALSE

```


<!--
ifelse(condition1,
       "Label1",           # used if condition1 is TRUE
       ifelse(condition2,
              "Label2",    # used if condition1 is FALSE and condition2 is TRUE
              "Label3"))   # used if BOTH condition1 and condition2 are FALSE
-->

---

## 4. Vectorized tests with ifelse() 

Nested `ifelse()` statements work, but they can make the code harder to read. 

Compare the previous code with this version that does not use them:

```{r, echo = TRUE, results = "hide"}

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

qualify %>%
  mutate(Status = "Rejected") %>%                                  # default label
  mutate(Status = ifelse(Scores >= 30, "Waitlisted", Status)) %>%  # update if condition met
  mutate(Status = ifelse(Scores > 35, "Admitted", Status))         # update if higher condition met
```

<!--
# or can also start from Admittd and down

qualify %>%
  mutate(Status = "Admitted") %>%                                   # start with highest category
  mutate(Status = ifelse(Scores <= 30, "Waitlisted", Status)) %>%  # downgrade if below admission threshold
  mutate(Status = ifelse(Scores < 25, "Rejected", Status))          # downgrade again if below waitlist threshold
-->



<br>

::: {.callout-tip title="Tip"}

If you have multiple conditions (more than two):

- Write multiple separate `mutate()` + `ifelse()` steps, as shown on this slide, rather than nesting them (as shown on the previous slide)

- When possible, use `case_when()` instead of `ifelse()` for better readability

:::

</br>


<!--
## When to Use `ifelse()` Instead of `case_when()`

While `case_when()` is better for multiple conditions, `ifelse()` is sometimes the better choice.

Use `ifelse()` when:

- You're checking **only one condition**
- You want a **simple inline transformation**  
  (e.g., `ifelse(x > 0, x * 2, x)`)
- You're assigning `NA` for some values  
  (e.g., `ifelse(x < 0, NA, x)`)

Use `case_when()` when:

- You have **3 or more conditions**
- You want better readability for **categorical recoding**

-->


---

## 4. Vectorized tests with ifelse() 

People often use `ifelse()` to recode variables in a dataframe. It's a good choice when the recoding can be done with a single `ifelse()`, as shown in this example.

Example: the variable `decisionDirection` takes four values (1 conservative, 2 liberal, 3 unspecifiable, and NA). Recode it to take three values (0 conservative, 1 liberal, and NA for both NA and unspecifiable)

```r
data %>%
  mutate(decisionDirection = ifelse(
    decisionDirection == 3, # condition to be evaluated
    NA,                     # if true, do NA
    decisionDirection - 1   # if false, subtract one to value
    ))
```

<!--
More on this:
https://docs.ycrc.yale.edu/r-novice-gapminder/07-control-flow/
Comparing ifelse() and for loop:
https://www.r-bloggers.com/2020/02/if-else-and-ifelse/
-->

---

## 4. Vectorized tests with ifelse() 

**Note there is also a tidyverse version of the base R `ifelse()` function!**

The tidyverse version is called `if_else()` <https://dplyr.tidyverse.org/reference/if_else.html>

The two are similar, hence I introduced only on here, pick the one you prefer.

---

## 💻 Practice

You can fine more practice exercises in today's class materials (downloaded from the website)

---

## Recap: What We Learned Today

- Control structures let us change the flow of execution in our code  
- We focused on conditional statements: how to make decisions using `if`, `else`, and `else if`
- We practiced different types of conditionals:
  - single test, multiple tests, and nested tests  
  - vectorized tests with `ifelse()` and `if_else()`


---

## To print these slides as pdf

Click on the icon bottom-right corner \> Tools \> PDF Export Mode \> Print as a Pdf

