---
title: "Introduction to R: Lecture 10"
subtitle: "Topics: Writing loops for iteration and repetition"
author: "Sabrina Nardin, Summer 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---

# Agenda {.center}

::: {.agenda-list}
1. For Loops
2. R Alternatives to For Loops
3. While Loops
:::

```{r, echo=FALSE, results='asis'}
cat('<span style="font-size: 0.8em; color: #666;">These slides were last updated on', format(Sys.Date(), '%B %d, %Y'), '</span>')
```

<!-- NOTES TO IMPROVE CURRENT LOOP SLIDES

Add a few slides that show the differences of looping over indexes or looping over elements and 
teach that concept form the slides; then leave the in-class demo code with the key and let students explore it in team and go around for questions. Total time 10 min slides and 10 go over code + 5 review

In the demo for loop: add code on data structure, e.g. show how to access columns of a df and their elements with the double and single square brackets 

Show difference of accessing single column in a df (that is a vector so use []) or elements of columns for which you need to use [[]]

When you teach for loops add the break and continue statements (currently not in slides!)

-->

---

# 1. For Loops {.slide .center .middle}

---

## What Are Control Structures?

Control structures let us write code that can:

- **Make decisions**  
  → with **conditional statements** (`if`, `if...else`, etc.)

- **Repeat actions**  
  → with **loops** (`for`, `while`)

Today we focus on loops!

---

## Definition of For Loops

For loops are the most common looping construct in many programming languages. They are used to **iterate over the elements of an object** (usually a list or vector) and perform an action on each one.

Syntax:
```
for (item in sequence of items) {
  action to be repeated for each item   
}
```

Example:
```r
for (item in c(1,2,3)) {
  print(item)
}
```

---

## Simple Example

```{r echo = TRUE, results = "hide"}

for (item in c(1,2,3)) {
  print(item)
}

```

Let's break this code down:

* the repeated action here is simple: `print(item)`

* `item` is a placeholder: its value changes with each loop iteration (it can be labeled any name you like)

* the number of times the statement block repeats depends on the number of items in the sequence — in this example: three items, thus three times

---

## Nested For Loop

For loops can be nested. In this case, the outer loop controls how many times the entire inner loop runs from start to finish: 

```{r echo = TRUE, results = "hide"}

for (i in c(1,2,3)) {
  print(i)
  for (j in c("cat", "dog", "squirrel", "rabbit")) {
    print(j)
  }
}
```

Let's break this code down:

- The outer loop `(i in 1:3)` runs 3 times

- For each execution of the outer loop, the inner loop runs 4 times

- That’s a total of 3 × 4 = 12 inner loop outputs, plus 3 outer loop prints

<!--
for (i in 1:3) {
  print(i)
  for (j in c('a', 'b')) {
    print(i)
    print(paste(i, "outer"))
    print(j)
    print(paste(i,j))
  }
}
-->

---

## For Loop Example #3

Observe this code. What does this code output in each run?

```{r echo = TRUE, results = "hide"}

for (i in c(1,2,3)) {
  print(i)
  print("Hello")
  sum <- 100 + 1
  print(sum)
}

```
<!-- What is this loop doing? Have someone describing it 
what happens if I add a print(i) at the end outside the loop: prints last i-->

---

## For Loop Example #4

This code takes the previous example, but defines the `sum` variable outside the loop (before it starts). What does this code output in each run?

```{r echo = TRUE, results = "hide"}

sum <- 0

for (i in c(1,2,3)) {
  print(i)
  print("Hello")
  sum <- sum + i
  print(sum)
}

```

---

## For Loop Example #4

What we just observed is a simple example of a common task: accumulating a running total. Use it when you are adding up scores, votes, or counts one at a time, and want to track the growing sum after each step

```{r echo = TRUE, results = "hide"}

sum <- 0

for (i in c(1,2,3)) {
  print(paste("Current i is:", i))
  print("Hello")
  sum <- sum + i
  print(paste("Current sum is:", sum))
}
```

---

## Save For Loop Output

The previous code correctly prints the updated sum (which changes in each loop) but it only saves the final one.

To keep all intermediate values, we need to save them in a vector:

```{r echo = TRUE, results = "hide"}

sum <- 0
output <- vector(mode = "integer", length = 3)

for (i in c(1,2,3)) {
  print(paste("Current i is:", i))
  print("Hello")
  sum <- sum + i
  print(paste("Current sum is:", sum))
  output[i] <- sum    # save the current sum at position i
}
output

```

Why we write `output[i]`? 

This code is good but not yet optimal. Let's see why.

---

## Looping over Indexes vs Looping over Elements 

```{r echo = TRUE, results = "hide"}

# Looping over elements
x <- c(2,4,3)
for (value in x) {
  print(value)
}

# Looping over indexes, can also use seq_along() 
x <- c(2,4,6)
for (i in 1:length(x)) {
  print(x[i])
}

# another slide with saving output vs printing, then go back to previous example and adjust also sum <- sum + i
```

::: {.callout-tip title="Tip: Prefer looping over indexes"}
Looping over indexes (i in 1:length(x)) gives you more control:

- You can access both the value and its position

- You can store, replace, or compare values more flexibly

- It's easier to save output or modify elements in-place

When you loop over values directly (e.g. for (value in x)):

You lose the index, so you can't easily:

store values by position

modify the original vector

track where you are in the loop

This matters if you want to update values or build a new vector


Best practice:
oop over indexes, not elements, when you need to:

Modify, save, or assign values by position (e.g., output[i] <- ...)

Ensure your loop works even if values repeat or are not valid indices

This gives you full control over both:

the position (i)

and the value at that position (x[i])

:::
---


## Compare Two Loops – What’s the Difference?

:::: {.columns}

::: {.column width="50%"}
### Version A

```{r echo = TRUE, results = "hide"}
sum <- 0
numbers <- c(1, 2, 3)
output <- vector(mode = "integer", 
                 length = length(numbers))

for (i in seq_along(numbers)) {
  print(paste("Current i is:", i))
  sum <- sum + i
  print(paste("Current sum is:", sum))
  output[i] <- sum
}
output
```
:::

::: {.column width="50%"}

### Version B

```{r echo = TRUE, results = "hide"}
sum <- 0
numbers <- c(1, 2, 3)
output <- vector(mode = "integer", 
                 length = length(numbers))

for (i in seq_along(numbers)) {
  print(paste("Current i is:", i))
  sum <- sum + numbers[i]
  print(paste("Current sum is:", sum))
  output[i] <- sum
}
output
```
:::

::::

---

## 💻 Practice

Compare Version A and Version B from the previous slide and identify their differences:

- Do they produce the same result the current `numbers` vector?
- Will they still behave the same if the values in `numbers` change?

**Try this:**
  - Change `numbers` to `c(5, 3, 10)` in both versions
  - Run both versions
  - Observe the outputs and explain the difference

---


## In sum

Summarize use of [i] for both output and sum


---

## 💻 Practice

You run a store and want to keep track of how many items were sold each day: you sold 5 items in day 1, 3 items on day 2, and 7 items on day 3

```r
daily_sales <- c(5, 3, 7)
```

Your Task:

- Use a for loop to go through each day’s sales 
- For each day, print something like: "Day 2 - Items sold: 3" (the message should change accordingly)
- Create an object called `total` to store a running total of all sales
- Create an object called `ouput` to store the total after each day (e.g. `output[2]` should store the total sales after day 2)

<!--
daily_sales <- c(5, 3, 7)

total <- 0
output <- vector(mode = "integer", length = length(daily_sales))

for (i in seq_along(daily_sales)) {
  print(paste("Day", i, "- Items sold:", daily_sales[i]))
  total <- total + daily_sales[i]
  output[i] <- total
}

-->


---

## Using For Loops with dataframes in R

Add a slide here on how loops work with dataframes or tibbles since we work a lot with them

---

Example with

---
 
Alternatives witout 


---

## 💻 Practice

---

## Recap: What We Learned Today

---

## Reminders

---

## To print these slides as pdf

Click on the icon bottom-right corner \> Tools \> PDF Export Mode \> Print as a Pdf
